\\Large Rising temperatures alter how and where mammals move

### 2025-10-14 writing/manuscript.Rmd
added habitat figure for caribou and wolves

\noindent We modeled each animal's movement using continuous-time movement models [@fleming_fine-scale_2014] via the `ctmm` package [v. 1.2.0, @fleming_ctmm_2023] for `R`. We then estimated each mammal's instantaneous speed at each observed location by applying the `ctmm::speeds()` function on all models with finite speed estimates [415 of 433$;$ @fleming_fine-scale_2014; @noonan_scale-insensitive_2019]. The remaining 18 movement models had sampling rates that were too coarse, relative to the animals' directional persistance, to reconstruct the animals' movement trajectories [@noonan_scale-insensitive_2019; @nathan_big-data_2022; @denicola_are_2025]. The models were for one mountain goat, 15 boreal caribou, and two southern mountain caribou (Table \@ref(tab:data-table)).

@@

@@

Considerations about changes in trophic interactions leads to another important caveat about the results we present. The estimated changes in movement behaviour cannot be fully attributed to physiological responses to temperature alone, since other aspects of habitats' seasonal phenology are (nonlinearly) correlated with temperature. For example, mountain goats' lower movement rates at temperatures near 0$^\circ$C in winter may be in part due to increased movement difficulty, since melting snow and rain-on-snow events increase the energetic costs of movement [@white_population_2025] and encounter rates with predators [@sullender_defining_2023]. However, milder temperatures can also increase plant growth, which allows goats to spend more time foraging within and less time travelling across patches [@charnov_optimal_1976; @white_mountain_2025]. Therefore, figure \@ref(fig:dist) illustrates an estimate of the total effects of temperature on each species' movement rates, but it does not explicitly account for energetic costs. Similarly, species' changes in habitat selection do not explicitly account for any changes in physiological or energetic costs. Additionally, forest cover, elevation, and distance from water are often correlated, so the estimated effects should be interpreted carefully, and they may not be applicable outside the study areaes used here. For example, both mountain goats and elk selected for higher elevation as temperatures warmed, but our models did not account for differences in forage availability at different elevations.

>>>>

\noindent We modeled each animal's movement using continuous-time movement models [@fleming_fine-scale_2014] via the `ctmm` package [v. 1.2.0, @fleming_ctmm_2023] for `R`. We then estimated each mammal's instantaneous speed at each observed location by applying the `ctmm::speeds()` function on all models with finite speed estimates [415 of 433$;$ @fleming_fine-scale_2014; @noonan_scale-insensitive_2019]. The remaining 18 movement models had sampling rates that were too coarse, relative to the animals' directional persistence, to reconstruct the animals' movement trajectories [@noonan_scale-insensitive_2019; @nathan_big-data_2022; @denicola_are_2025]. The models were for one mountain goat, 15 boreal caribou, and two southern mountain caribou (Table \@ref(tab:data-table)).

@@

The three habitat variables were nonlinearly correlated. Within the boreal caribou range (Fig. \@ref(fig:boreal-rss)A), locations with an elevation of 0.9-1.1 km had a median forest cover of 100% and a median distance to water of 1.8 km. In contrast, elevations above 1750 m did not have forest cover above 90% and had a median distance to water of 1.5 km. Local predictions of RSS revealed that boreal caribou and wolves selected for similar habitats (Fig. \@ref(fig:boreal-rss)B), but caribou tended to avoid the river system that wolves strongly selected for. Both species selected for the same loop of low-elevation habitat (< 500 m), but their selection strength for the area varied with temperature. For caribou, selection for the area was stronger at -20$^\circ$C and weaker at warmer temperatures, while, for wolves, selection was stronger at +20$^\circ$C. Both species' HRSFs indicated selection against elevations approximately between 0.5 km and 0.9 km but selection for elevations approximately above 0.9 km. After data cleaning and the removal of non-range-resident animals, the highest observed elevations were 1 km for caribou and 1.733 km for wolves. Quadrature points had maximum elevations of 1.1 km for caribou and 1.3 km for wolves.

```{r, eval=FALSE, echo=FALSE}
median(filter(resources, elevation_m > 900, elevation_m < 1100)$forest_perc)
median(filter(resources, elevation_m > 900, elevation_m < 1100)$dist_water_m)

max(filter(resources, forest_perc > 90)$elevation_m)
median(filter(resources, elevation_m > 1750)$dist_water_m)

ggplot(resources, aes(elevation_m, forest_perc)) + geom_hex(bins = 20)
ggplot(resources, aes(elevation_m, dist_water_m)) + geom_hex(bins = 20)
ggplot(resources, aes(dist_water_m, forest_perc)) + geom_hex(bins = 20)

e_rast <- rast('data/resource-rasters/bc-buffered-dem-z6.tif') %>%
    crop(st_buffer(st_transform(bounds, crs(.)), 1e5)) %>%
    project(f_rast) %>%
    crop(bounds)

plot(e_rast < 500)
plot(e_rast < 900)
plot(e_rast < 1000)

d %>%
  filter(species %in% c('Rangifer tarandus boreal', 'Canis lupus')) %>%
  summarize(elevation_m = max(elevation_m), .by = c(species, detected))
```

```{r boreal-rss, fig.cap="\\textbf{Temperature affects boreal caribou's and wolves' habitat selection and, consequently, their encounter rates.} Panel A shows the rasters of the three habitat variables used in the hierarchical resource selection functions, while panel B shows the relative selection strength (RSS) for boreal caribou and wolves, as a function of temperature. RSS values were re-centered and capped to between 0.25 and 4 ($2^{\\pm 2}$) to improve readability."}
knitr::include_graphics('../figures/boreal-caribou-wolf-habitat.png')
```

@@

Considerations about changes in trophic interactions leads to another important caveat about the results we present. The estimated changes in movement behaviour cannot be fully attributed to physiological responses to temperature alone, since other aspects of habitats' seasonal phenology are (nonlinearly) correlated with temperature. For example, mountain goats' lower movement rates at temperatures near 0$^\circ$C in winter may be in part due to increased movement difficulty, since melting snow and rain-on-snow events increase the energetic costs of movement [@white_population_2025] and encounter rates with predators [@sullender_defining_2023]. However, milder temperatures can also increase plant growth, which allows goats to spend more time foraging within and less time travelling across patches [@charnov_optimal_1976; @white_mountain_2025]. Therefore, figure \@ref(fig:dist) illustrates an estimate of the total effects of temperature on each species' movement rates, but it does not explicitly account for energetic costs. Similarly, species' changes in habitat selection do not explicitly account for any changes in physiological or energetic costs. Additionally, forest cover, elevation, and distance from water are often correlated, so the estimated effects should be interpreted carefully, and they may not be applicable outside the study areas used here. For example, both mountain goats and elk selected for higher elevation as temperatures warmed, but our models did not account for differences in forage availability at different elevations.
